Addition and removal energies of circular quantum dots

### 2017-06-20 Manuscript/paper.tex
Fixes in response to Nathan's review

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Samuel J. Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{Scott Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract}
  We present and compare several many-body methods as applied to two-dimensional quantum dots with circular symmetry.  We calculate the approximate ground state energy using a harmonic oscillator basis optimized by Hartree--Fock (HF) theory and further improve the ground state energy using two post-HF methods: in-medium similarity renormalization group (IM-SRG) and coupled cluster singles-and-doubles (CCSD).  With the application of quasidegenerate perturbation theory (QDPT) or the equation-of-motion (EOM) method, we obtain addition and removal energies as well.  Our results are benchmarked against full configuration interaction (FCI) and diffusion Monte Carlo (DMC) where available.  We examine the rate of convergence and perform extrapolations to the infinite basis limit using a power-law model.
\end{abstract}

@@

The behavior of strongly confined electrons is of fundamental concern to many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificially constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as they lie at a critical point where both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

@@

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method, M\o ller-Plesset perturbation theory (MP), the in-medium similarity renormalization group (IM-SRG) method, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT), and equation-of-motion (EOM) methods.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) theory \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results, where available, from the existing literature.

Of these methods, IM-SRG is a recently developed technique that has shown significant promises.  Similarity renormalization group (SRG) methods are a family of methods that transform the Hamiltonian into a band- or block-diagonal form through a continuous sequence of unitary transformations.  The goal of such a transformation is to reduce the coupling between a subspace of interest -- such as the ground state or a set of low-lying states -- and the remaining Hilbert space.  It has successfully been applied to systems with various underlying potentials to calculate their binding energy and other observables, especially in nuclear theory \cite{ScottSRG,PhysRevC.75.061001,SRGThreeDim}.  The \emph{in-medium} SRG (IM-SRG) is an adaptation of the SRG method to evolve Hamiltonians in a truncated Fock space, which avoids the prohibitive cost of constructing a full many-body Hamiltonian.  The effect of the truncation is minimized through the normal-ordering technique, which re-expands the operators with respect to a nearby many-body reference state rather than the far-away vacuum state.

The article is organized as follows: Section \ref{sec:formalism} introduces first (\ref{subsec:modelHamiltonian}) the Hamiltonian we use to model circular quantum dots, and gives afterwards an overview of the \textit{ab initio} many-body methods used in this paper: Hartree--Fock, IM-SRG, CCSD, QDPT, EOM.  Our results are presented in Section \ref{sec:results}.  We analyze and compare the differences between the various methods and also between different quantum dot systems.  We examine the usefulness of extrapolation techniques in improving the precision of the results.  Section \ref{sec:conclusions} concludes our work and gives perspectives for future work.

@@

We shall model the circular quantum dot system as a collection of $N$ nonrelativistic electrons of mass $m$ in two-dimensional space, trapped by an external harmonic-oscillator potential of the form $m \omega^2r^2 / 2$, where $\omega$ is the angular frequency of the oscillator trap and $r$ is the distance from the center of the trap.  The electrons interact with each other through the standard Coulomb interaction $e^2 / (4 \pi \epsilon R)$, where $e$ is the charge of each electron, $\epsilon$ is the permittivity of the medium, and $R$ is the distance between the two interacting electrons.

@@

The behavior of each electron in isolation is described by the nonrelativistic kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2, &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2, &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r}).
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}.
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H_1 + \hat H_2,
\end{align*}
where $\hat{H}_1$ and $\hat{H}_2$ are respectively the one- and two-body parts of the Hamiltonian, namely
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}_1 &\equiv \sum_{\alpha = 1}^N h(\hat{\bm p}_\alpha, \hat{\bm r}_\alpha), &
\hat{H}_2 &\equiv \sum_{\alpha = 1}^N \sum_{\beta = 1}^{\alpha - 1} v(\hat{\bm r}_\alpha, \hat{\bm r}_\beta).
\end{align}
The operator $\hat{\bm r}_\alpha$ is the position operator of the $\alpha$-th particle,\footnote{The ordering of the particle labels $\alpha$ is unimportant as they exist only for the purpose of bookkeeping.} and $\hat{\bm p}_\alpha \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_\alpha}$ is its momentum operator.

@@

The noninteracting part of the many-body Hamiltonian $\hat{H}_1$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2.
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle,
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock-Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \frac{\partial}{\partial \varphi}$.

@@

The energy of a Slater determinant is given by a sum over the occupied single-particle energies,
\begin{align*}
  \hat{H}_1 | p_1 \ldots p_N \rangle = (\varepsilon_{p_1} + \cdots + \varepsilon_{p_N}) | p_1 \ldots p_N \rangle.
\end{align*}
It is possible for Slater determinants occupying different sets of single-particle states to be degenerate in energy.  This is possible even for ground states of $\hat{H}_1$.  However, if the number of particles $N$ satisfies $N = K_{\mathrm{F}} (K_{\mathrm{F}} + 1)$ for some nonnegative integer $K_{\mathrm{F}}$, there would be just enough particles to form a closed-shell Slater determinant, leading to a unique, well-isolated ground state.  The values of $N$ at which this occurs are said to be \textit{magic numbers}.  We call $K_{\mathrm{F}}$ the \textit{number of filled shells} (or ``Fermi level'').  In particular, a single-particle state is occupied in the ground state Slater determinant if and only $k < K_{\mathrm{F}}$, where $k$ is the shell index of the single-particle state as defined in \eqref{eq:shell_index}.

@@

To perform numerical calculations, we further assume that each unknown state $|p'\rangle$ is built from a linear combination of known states $|p\rangle$, with an unknown matrix of coefficients $\bm C$ defined via the transformation equation
\begin{align*}
  |p'\rangle \equiv \sum_p |p\rangle C_{p p'}.
\end{align*}
This allows the problem to be reduced from an abstract minimization problem \eqref{eq:hfexpectation} to a concrete numerical problem.  The caveat is that the set of known functions be large enough to capture the relevant behavior of the system.

@@

The goal then is to find the coefficients $\bm C$ that minimize the energy $E_{\Phi}$,
\begin{align}
  E_{\Phi} &= \sum_{i'} \langle i' | \hat{H}_1 | i' \rangle + \sum_{i' j'} \langle i' j' | \hat{H}_2 | i' j' \rangle \label{eq:hfenergy},
\end{align}
where
\begin{align}
  \langle p' | \hat{H}_1 | q' \rangle &= \sum_{p q} C_{p p'}^* \langle p | \hat{H}_1 | q \rangle C_{q q'}^{}, \label{eq:hftransform1} \\
  \langle p' q' | \hat{H}_2 | r' s' \rangle &= \sum_{p q r s} C_{p p'}^* C_{q q'}^* \langle p q | \hat{H}_2 | r s \rangle C_{r r'}^{} C_{s s'}^{}, \label{eq:hftransform2}
\end{align}

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree--Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm{F} \bm{C} = \bm{C} \bm{\varepsilon},
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p q} \equiv \langle p | \hat{H}_1 | q \rangle + \sum_{r s i'} C_{r i'}^* \langle p r | \hat{H}_2 | q s \rangle C_{s i'}^{},
\end{align}
and $\bm{\varepsilon}$ is a vector of Lagrange multipliers: each element $\varepsilon_{p'}$ is associated with a specific single-particle state $|p'\rangle$.

Aside from trivial cases, the HF equation is generally solved using an iterative algorithm.  We begin with an initial guess $\bm{C}^{(j)}$ on the $j$-th iteration, which is fed into Eq.\ \eqref{eq:fock} to produce the Fock matrix.  This is then used in Eq.\ \eqref{eq:hartreefock}, which leads to a standard eigenvalue problem from which $\bm{C}^{(j + 1)}$ arise as the matrix of eigenvectors and $\bm{\varepsilon}^{(j + 1)}$ as the vector of eigenvalues.  This process can be repeated indefinitely until $\bm{C}$ approaches a fixed point.  While in theory it is possible for the solution to never reach a fixed point, or that it may require a unfeasibly large number of iterations, in practice this naive approach can adequately provide solutions for many cases.  In other cases where it is insufficient, methods such as DIIS, Broyden's method, or even \textit{ad hoc} linear mixing can improve and accelerate convergence greatly.  Therefore, the possibility of slow or non-convergence is generally not a concern in practice.

@@

Since HF restricts the ground state to merely a single Slater determinant of single-particle states, it cannot provide an exact solution to a problem where multi-particle correlations are present even if the single-particle basis is not truncated (infinite in size).  The discrepancy between the Hartree--Fock energy and the exact ground state energy is often referred to as the \textit{correlation energy}, by definition.  The focus of post-HF methods such as IM-SRG or coupled cluster is to add corrections beyond mean-field approximations such as HF.

To make use of the HF solution as the reference state for post-HF calculations, we transform the matrix elements via Eqs.\ \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  In effect, this means we are no longer operating within the harmonic oscillator single-particle basis, but a HF-optimized single-particle basis.  However, we will omit the prime symbols as the post-HF methods are generic and can be used in any basis, whether optimized by HF or not.

@@

Note that the loss of three- and higher-body terms (\textit{operator truncation}) is only one out of the two sources of error in this method.  The other source of error is due to the \textit{basis truncation}, a concern for any method that relies on a finite single-particle basis such as IM-SRG.  The latter can be reduced by increasing the size of the basis at the expense of greater computational effort, albeit the cost increases much less rapidly in this direction.  In IM-SRG the CPU cost is polynomial with respect to $N_{\mathrm{b}}$, the number of states in the single-particle basis.  For IM-SRG(2) in particular, the CPU cost scales roughly as $\mathcal{O}(N_{\mathrm{b}}^6)$.

@@

Through the flow equation and with an appropriate choice of the generator $\hat{\eta}$, the evolved state $\hat U(s) \ket{\Phi}$ will gradually approach a more ``diagonal'' form.  If the ``diagonal'' form decouples the ground state from the excited states, then $\hat U(\infty) \ket{\Phi}$ would yield the exact ground state solution of the problem if no operator or basis truncations are made.  In particular, $H_\varnothing(\infty)$ would be the exact ground state energy.

@@

Apart from Wegner's generator, there exist several other ones in literature. One choice, proposed by White \cite{White:cond-mat0201346}, makes numerical approaches much more efficient.  The problem with the Wegner generator is the widely varying decaying speeds of the Hamiltonian matrix elements.  Terms with large energy separations from the ground state are suppressed initially, followed by those with smaller energy separations.  This leads to stiffness in the flow equation, leading to numerical difficulties in solving the set of coupled differential equations.

The White generator takes an alternative approach, which is well suited for problems where one is mainly interested in the ground state of a system.  Firstly, instead of driving all off-diagonal elements of the Hamiltonian to zero, the generator focuses exclusively on those ones that are coupled to the reference state $\Phi$ so as to decouple the reference state from the remaining Hamiltonian.  This reduces the amount of change done to the Hamiltonian, reducing the accuracy lost from the operator truncation.  Secondly, the rate of decay in Hamiltonian matrix elements are approximately normalized by dividing the generator matrix elements by an appropriate factor.  This ensures that the affected elements decay at approximately the same rate, reducing the stiffness of the flow equations.

The White generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta'_{a i} &\equiv \frac{H_{a i}}{\tilde{\Delta}_{a i}}, &
\eta'_{a b i j} &\equiv \frac{H_{a b i j}}{\tilde{\Delta}_{a b i j}},
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}_{a i} &\equiv \Delta_{a i} - H_{a i a i}, \notag \\
\tilde{\Delta}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H_{p_i p_i} -  H_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H_{a b a b} - H_{a i a i} - H_{b i b i} \notag \\
  &\qquad + H_{i j i j} - H_{a j a j} - H_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

\subsection{Coupled-cluster theory}
\label{subsec:cctheory}

>>>>

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Samuel J. Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{The Ohio State University, Columbus, OH 43210, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{Scott K. Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract}
  We present and compare several many-body methods as applied to two-dimensional quantum dots with circular symmetry.  We calculate the approximate ground state energy using a harmonic oscillator basis optimized by Hartree--Fock (HF) theory and further improve the ground state energy using two post-HF methods: in-medium similarity renormalization group (IM-SRG) and coupled cluster with singles and doubles (CCSD).  With the application of quasidegenerate perturbation theory (QDPT) or the equations-of-motion (EOM) method to the results of the previous two methods, we obtain addition and removal energies as well.  Our results are benchmarked against full configuration interaction (FCI) and diffusion Monte Carlo (DMC) where available.  We examine the rate of convergence and perform extrapolations to the infinite basis limit using a power-law model.
\end{abstract}

@@

The behavior of strongly confined electrons is a major problem for many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificially constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as they lie at a critical point where both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

@@

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method, M\o ller-Plesset perturbation theory (MP), the in-medium similarity renormalization group (IM-SRG) method, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT), and equations-of-motion (EOM) methods.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) theory \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results, where available, from the existing literature.

Of these methods, IM-SRG is a recently developed technique that has shown significant promises.  Similarity renormalization group (SRG) methods are a family of methods that transform the Hamiltonian into a band- or block-diagonal form through a continuous sequence of unitary transformations.  The goal of such a transformation is to reduce the coupling between a subspace of interest -- such as the ground state or a set of low-lying states -- and the remaining Hilbert space.  It has successfully been applied to systems with various underlying potentials to calculate their binding energy and other observables, especially in nuclear theory \cite{ScottSRG,PhysRevC.75.061001,SRGThreeDim}.  The \emph{in-medium} SRG (IM-SRG) method adapts the SRG approach to evolve Hamiltonians in a truncated Fock space that is centered around an approximate reference state rather than the physical vacuum, which substantially reduces the importance of the computationally costly higher-body operators.

This article is organized as follows: Section \ref{sec:formalism} introduces first (\ref{subsec:modelHamiltonian}) the Hamiltonian we use to model circular quantum dots, and gives afterwards an overview of the \textit{ab initio} many-body methods used in this paper: HF, IM-SRG, CCSD, QDPT, EOM.  Our results are presented in Section \ref{sec:results}.  We analyze and compare the differences between the various methods and also between different quantum dot systems.  We examine the utility of extrapolation techniques in improving the precision of our results.  Section \ref{sec:conclusions} concludes our work and gives perspectives for future work.

@@

We shall model the circular quantum dot system as a collection of $N$ nonrelativistic electrons of mass $m$ in two-dimensional space, trapped by an external harmonic-oscillator potential of the form $m \omega^2r^2 / 2$, where $\omega$ is the angular frequency of the oscillator trap and $r$ is the distance from the center of the trap.  The electrons interact with each other through the standard Coulomb interaction $e^2 / (4 \pi \epsilon R)$, where $e$ is the electron charge, $\epsilon$ is the permittivity of the medium, and $R$ is the distance between the two interacting electrons.

@@

The behavior of each electron in isolation is described by the nonrelativistic kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2, &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2, &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r}).
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}.
\end{align*}
The full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H_1 + \hat H_2,
\end{align*}
where $\hat{H}_1$ and $\hat{H}_2$ are respectively the one- and two-body parts of the Hamiltonian, namely
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}_1 &\equiv \sum_{\alpha = 1}^N h(\hat{\bm p}_\alpha, \hat{\bm r}_\alpha), &
\hat{H}_2 &\equiv \sum_{\alpha = 1}^N \sum_{\beta = 1}^{\alpha - 1} v(\hat{\bm r}_\alpha, \hat{\bm r}_\beta).
\end{align}
The operator $\hat{\bm r}_\alpha$ is the position operator of the $\alpha$-th particle,\footnote{The ordering of the particle labels $\alpha$ is unimportant as they exist only for the purpose of bookkeeping.} and $\hat{\bm p}_\alpha \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_\alpha}$ is its momentum operator.

@@

The noninteracting part of the many-body Hamiltonian $\hat{H}_1$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2.
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle time-independent Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle,
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock-Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \frac{\partial}{\partial \varphi}$.

@@

The energy of a Slater determinant is given by a sum over the occupied single-particle energies,
\begin{align*}
  \hat{H}_1 | p_1 \ldots p_N \rangle = (\varepsilon_{p_1} + \cdots + \varepsilon_{p_N}) | p_1 \ldots p_N \rangle.
\end{align*}
It is possible for Slater determinants occupying different sets of single-particle states to be degenerate in energy.  This is possible even for ground states of $\hat{H}_1$.  However, if the number of particles $N$ satisfies $N = K_{\mathrm{F}} (K_{\mathrm{F}} + 1)$ for some nonnegative integer $K_{\mathrm{F}}$, there would be just enough particles to form a closed-shell Slater determinant, leading to a unique, well-isolated ground state.  The values of $N$ at which this occurs are said to be \textit{magic numbers}.  We call $K_{\mathrm{F}}$ the \textit{number of filled shells} (or ``Fermi level'').  In particular, a single-particle state is occupied in the ground state Slater determinant if and only if $k < K_{\mathrm{F}}$, where $k$ is the shell index of the single-particle state as defined in \eqref{eq:shell_index}.

@@

To perform numerical calculations, we further assume that each unknown state $|p'\rangle$ is built from a linear combination of known states $|p\rangle$, with an unknown matrix of coefficients $\bm C$ defined via the transformation equation
\begin{align*}
  |p'\rangle \equiv \sum_p |p\rangle C_{p p'}.
\end{align*}
This allows the problem to be reduced from an abstract minimization problem \eqref{eq:hfexpectation} to a concrete numerical problem.  The caveat is that the set of known functions must be large enough to capture the relevant behavior of the system.

@@

The goal then is to find the coefficients $\bm C$ that minimize the \textit{Hartree--Fock energy} $E_{\Phi}$,
\begin{align}
  E_{\Phi} &= \sum_{i'} \langle i' | \hat{H}_1 | i' \rangle + \sum_{i' j'} \langle i' j' | \hat{H}_2 | i' j' \rangle \label{eq:hfenergy},
\end{align}
where
\begin{align}
  \langle p' | \hat{H}_1 | q' \rangle &= \sum_{p q} C_{p p'}^* \langle p | \hat{H}_1 | q \rangle C_{q q'}^{}, \label{eq:hftransform1} \\
  \langle p' q' | \hat{H}_2 | r' s' \rangle &= \sum_{p q r s} C_{p p'}^* C_{q q'}^* \langle p q | \hat{H}_2 | r s \rangle C_{r r'}^{} C_{s s'}^{}, \label{eq:hftransform2}
\end{align}

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the self-consistent \textit{Hartree--Fock equations}:
\begin{align} \label{eq:hartreefock}
  \bm{F} \bm{C} = \bm{C} \bm{\varepsilon},
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p q} \equiv \langle p | \hat{H}_1 | q \rangle + \sum_{r s i'} C_{r i'}^* \langle p r | \hat{H}_2 | q s \rangle C_{s i'}^{},
\end{align}
and $\bm{\varepsilon}$ is a vector of Lagrange multipliers, which serve to constrain the orthonormality of the single-particle basis.  Each multiplier $\varepsilon_{p'}$ is associated with a specific single-particle state $|p'\rangle$.

Aside from trivial cases that are analytically solvable, the HF equation is generally solved numerically using an iterative algorithm.  We begin with an initial guess $\bm{C}^{(j)}$ on the $j$-th iteration, which is fed into Eq.\ \eqref{eq:fock} to produce the Fock matrix.  This is then used in Eq.\ \eqref{eq:hartreefock}, which leads to a standard eigenvalue problem from which $\bm{C}^{(j + 1)}$ arise as the matrix of eigenvectors and $\bm{\varepsilon}^{(j + 1)}$ as the vector of eigenvalues.  This process can be repeated indefinitely until $\bm{C}$ approaches a fixed point (self-consistency).  While in theory it is possible for the solution to never reach a fixed point, or that it may require a unfeasibly large number of iterations, in practice this naive approach can adequately provide solutions for many cases.  In other cases where it is insufficient, methods such as DIIS \cite{PULAY1980393,JCC:JCC540030413}, Broyden's method \cite{broyden1965class}, or even \textit{ad hoc} linear mixing can improve and accelerate convergence greatly.  Therefore, the possibility of slow or non-convergence is generally not a concern in practice.

@@

Since HF restricts the ground state to merely a single Slater determinant of single-particle states, it cannot provide an exact solution to a problem where multi-particle correlations are present even if the single-particle basis is not truncated (infinite in size).  The discrepancy between the HF energy and the exact ground state energy is often referred to as the \textit{correlation energy}, by definition.  The focus of post-HF methods such as IM-SRG or CC is to add corrections beyond mean-field approximations such as HF.

To make use of the HF solution as the reference state for post-HF calculations, we transform the matrix elements via Eqs.\ \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  In effect, this means we are no longer operating within the harmonic oscillator single-particle basis, but rather a HF-optimized single-particle basis.  However, we will omit the prime symbols as the post-HF methods are generic and can be used in any basis, whether optimized by HF or not.

@@

Note that the loss of three- and higher-body terms (\textit{operator truncation}) is only one out of the two sources of error in this method.  The other source of error is due to the \textit{basis truncation}, a concern for any approach that relies on a finite single-particle basis, including HF, IM-SRG, CC, and many others.  This second source of error can be reduced by increasing the size of the basis at the expense of greater computational effort, albeit the cost increases much less rapidly in this direction.  The CPU cost of IM-SRG methods is polynomial with respect to $N_{\mathrm{b}}$, the number of states in the single-particle basis.  For IM-SRG(2) in particular, the CPU cost scales roughly as $\mathcal{O}(N_{\mathrm{b}}^6)$.

@@

With an appropriate choice of the generator $\hat{\eta}$, the evolved state $\hat U(s) \ket{\Phi}$ will gradually approach a more ``diagonal'' form.  If the ``diagonal'' form decouples the ground state from the excited states, then $\hat U(\infty) \ket{\Phi}$ would yield the exact ground state solution of the problem if no operator or basis truncations are made.  In particular, $H_\varnothing(\infty)$ would be the exact ground state energy.

@@

There exist several other generators in literature.  One choice, proposed by White \cite{White:cond-mat0201346}, makes numerical approaches much more efficient.  The problem with the Wegner generator is the widely varying decaying speeds of the Hamiltonian matrix elements.  Terms with large energy separations from the ground state are suppressed initially, followed by those with smaller energy separations.  This leads to stiffness in the flow equation, which in turn causes numerical difficulties when solving the set of coupled differential equations.

The White generator takes an alternative approach, which is well suited for problems where one is mainly interested in the ground state of a system.  Firstly, instead of driving all off-diagonal elements of the Hamiltonian to zero, the generator focuses exclusively on those that are coupled to the reference state $\Phi$ so as to decouple the reference state from the remaining Hamiltonian.  This reduces the amount of change done to the Hamiltonian, reducing the accuracy lost from the operator truncation.  Secondly, the rate of decay in Hamiltonian matrix elements are approximately normalized by dividing the generator matrix elements by an appropriate factor.  This ensures that the affected elements decay at approximately the same rate, reducing the stiffness of the flow equations.

The White generator is explicitly constructed in the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta'_{a i} &\equiv \frac{H_{a i}}{\tilde{\Delta}_{a i}}, &
\eta'_{a b i j} &\equiv \frac{H_{a b i j}}{\tilde{\Delta}_{a b i j}},
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}_{a i} &\equiv \Delta_{a i} - H_{a i a i}, \notag \\
\tilde{\Delta}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H_{p_i p_i} -  H_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H_{a b a b} - H_{a i a i} - H_{b i b i} \notag \\
  &\qquad + H_{i j i j} - H_{a j a j} - H_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

\subsection{Coupled cluster theory}
\label{subsec:cctheory}

Coupled cluster (CC) theory is based on expressing the $N$-particle correlated wave function $\ket{\Psi}$ using the exponential ansatz,
\begin{align*}
  \ket{\Psi} = e^{\hat{T}} \ket{\Phi},
\end{align*}
where $\ket{\Phi}$ is the reference state as before.  The cluster operator $\hat{T} \equiv \hat{T}_1 + \hat{T}_2 + \cdots + \hat{T}_N$, is composed of $k$-particle $k$-hole excitation operators, $\hat{T}_k$,
\begin{align} \label{eq:cc_amps}
  \hat{T}_k \equiv \left(\frac{1}{k!}\right)^2 \sum_{\substack{a_1 \ldots a_k \\ i_1 \ldots i_k}} t_{a_1 \ldots a_k i_1 \ldots i_k} \normord{\hat{a}_{a_1}^\dagger \ldots \hat{a}_{a_k}^\dagger \hat a_{i_k}^{} \ldots \hat{a}_{i_1}^{}},
\end{align}
where the unknown matrix elements, $t_{a_1 \ldots a_k i_1 \ldots i_k}$, are known as \textit{cluster amplitudes} \cite{shavitt2009many}.
